---
title: "Assignment 2: Coding Basics"
author: "Fiona Kelley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. The name "x" is assigned to a sequence from 1 to 30, increasing by threes. 
x<-seq(1,30,3)

#2. The names "x_mean" and "x_median" are assigned to the mean and median values computed. 
x_mean<-mean(x)
x_median<-median(x)

#3. The mean and median values are being compared to determine if the mean is greater than the median. 
x_mean>x_median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#Vectors are defined for names of students, test scores values, and whether the students have passed the test.  

#student_name is a character vector. 
student_name = c("Fiona", "Declan", "Aidan", "Maureen")
#test_scores is a numerical vector. 
test_scores = c(95,48,74,92)
#passed_test is a logical vector. 
passed_test = c(TRUE, FALSE, TRUE, TRUE)

#Vectors are placed within a data frame that prints the values in a list format. 
student_information<-list(
student_name,
test_scores, 
passed_test
)

#I utilized Chat GPT to assist with generating the code for this section. 

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A single data frame can contain different types of vectors, such as the data frame student_information contains character, numerical, and logical vectors. On the contrary, only one uniform vector type can be contained across a matrix.  

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#This function uses an if/else statement to evaluate student test scores and determine whether these are passing grades equal to or above 50. 
test_scores_passing_check<-function(test_scores){
  ifelse(test_scores>=50, "Pass", "Fail")
}

#This then runs the function for test scores. 
test_scores_passing_check(test_scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The 'ifelse' option worked, as opposed to 'if' and 'else', because the latter option produced an error regarding it's inability to evaluate vectors.
